A Bayesian ranking statistic to find high-mass black holes in LIGO--Virgo data

### 2021-03-29 main.tex
Updates from Overleaf

One proposed heave-seed pathway involves the direct collapse of massive, metal-free Population III stars. Typically, massive stars $> 100 \ \msun$ can eject more than half of their mass in the form of radiatively driven stellar winds~\cite{}. However, at zero metallicity, mass loss through stellar winds is negligible \cite{Kudritzki 2000}. Consequently,  Population III stars have the opportunity to complete their life-cycles without losing a significant portion of their mass. Moreover, Population III stars with masses above $230 \ \msun$ will avoid electron-positron pair-instability. Therefore, oxygen and silicon burning will not drive an explosion, allowing the star to undergo a complete collapse to leave behind an intermediate-mass black hole remnant with a mass of $120-280 \ \msun$, dependant on the initial star mass \cite{Bond et al. 1984}. As the fraction of Population III stars with masses above $230 \ \msun$ is uncertain, the number of expected intermediate-mass black holes from these stars is also ambiguous~\cite{Coleman Miller, Formation Scenarios for Intermediate-Mass Black Holes}.

@@

\rs{\sout{It is possible to measure the mass of nearby supermassive black holes by directly observing the dynamic motion of stars nearby the supermassive black hole (such as Sgr A$^*$~). However, this dynamic measurement method is challenging to utilise for  intermediate-mass black holes as their sphere of influence is too small to resolve motions~\cite{}. Additionally, dynamic-measurements that have alluded to  intermediate-mass candidates, e.g. the candidate at the centre of the globular cluster 47 Tucanae~\cite{kiziltan 2017}, can also be attributed to simpler models without a central  intermediate-mass black hole~\cite{freire 2017}. }}

@@

Although a significant fraction of the glitches can be removed by testing them for coherence amongst various ground based detectors and performing matched-filtering, these methods are insufficient to remove all the glitches. One method to account for more glitches while searching for high total mass CBC gravitational-waves is by utilising the astrophysical Bayesian odds~\cite{bci, BCR1, BCR2, bcr_gw151216, bayesian_odds}. A true Bayesian odds, calculated without using bootstrap techniques, can provide more accurate estimate of significance that does note depend on the search pipeline~\cite{BCR2, bcr_gw151216,  bayesian_odds}. In this paper, we utilise a Bayesian method, called the Bayesian Coherence Ratio \bcr~\cite{BCR1}, to rank the candidate gravitational-wave signals from high-mass compact binary coalescences (systems with total masses in the range of $55-500\ \msun$) in the detector data recorded during O2. Although the \bcr, utilising bootstrap techniques, does not provide the true Bayesian odds,  it utilizes Bayesian evidences which describe the explicit probability of data under the hypothesis that it contains coherent signals vs incoherent glitches.

@@

The standard framework to identify CBC gravitational-wave signals hidden in data is by quantifying the significance of candidates with null-hypothesis significance testing. In this framework, the candidates' ranking statistic is compared against a background distribution in a frequentist approach. \rs{add a sentence describing what the ranking stats are for the different pipelines (key words: matched filtering, SNR etc...) and add citations!} On the other hand, the standard framework for performing parameter estimation and model selection in gravitational-wave astronomy is Bayesian inference. This work utilises Bayesian inference to calculate the Bayesian Coherence odds-ratio~\cite{BCR1}, \bcr, of high-mass candidates in LIGO's second observing run. We use the \bcr not as an odds-ratio but instead as a ranking statistic, a step toward building a unified Bayesian framework to search for candidates and estimate their parameters.

@@

Bayes theorem states that the posterior probability distribution $p(\parameters|d,\mathcal{H})$ for data $d$ and a vector of parameters \parameters that describe a model which quantifies a hypothesis $\mathcal{H}$, is given by
\begin{equation}
p(\parameters|d,\mathcal{H}) = \frac{\mathcal{L}(d|\parameters, \mathcal{H}) \ \pi(\parameters | \mathcal{H})}{\mathcal{Z}(d|\mathcal{H})}\ , 
\end{equation}
where $\mathcal{L}(d|\parameters, \mathcal{H})$ is the likelihood of the data given the parameters \parameters and the hypothesis, $\pi(\parameters | \mathcal{H})$ is the prior probability of the parameters, and finally,
\begin{equation}
    \mathcal{Z}(d|\mathcal{H}) = \int\limits_{\parameters} \ \mathcal{L}(d|\parameters,\mathcal{H}) \ \pi(\parameters | \mathcal{H}) \ d\parameters
\end{equation} is the likelihood after marginalising over the parameters \parameters.  To compare two hypotheses $\mathcal{H}_A$ and $\mathcal{H}_B$ with the Bayes theorem one can calculate an odds-ratio
\begin{equation}
    \mathcal{O}^A_B = \frac{\mathcal{Z}^A\ \pi(\parameters^A)}{\mathcal{Z}^B\ \pi(\parameters^B)}\ ,
\end{equation}
where $\mathcal{Z}^A$ and $\mathcal{Z}^B$ are the shorthand for the evidences  $\mathcal{Z}(d|\mathcal{H}_A)$ and $\mathcal{Z}(d|\mathcal{H}_B)$. The odds-ratio can tell us which of the two hypotheses is more likely. For example, if $\mathcal{O}^A_B >> 1$, then this odds ratio indicates that the $\mathcal{H}_A$ describes the data much better than $\mathcal{H}_B$.

@@

For our study, we filter the background, simulated and candidate events to include only high-mass events with masses in the ranges of the parameters presented in Table~\ref{tab:parameters}. A plot of the \pycbc triggers from one time-frame, during April 23 - May 8, 2017, is presented in Figure~\ref{fig:templateBank}. This figure also depicts the gravitational-wave templates used during the search through this time-frame of data.

@@

\subsection{Calculating the BCR for triggers}
\avi{Sat 13th: Just realised that some parts of this dont make sense without the material present in Apdx~\ref{sec:bayesianEvidEval}}...
To evaluate $Z^S$, $Z^G_i$ and $Z^N_i$ and calculate the \bcr Eq.~\ref{eq:bcr} for triggers, we carry out Bayesian inference with \bilby~\cite{bilby, bilby_pipe}, employing \dynesty~\cite{dynesty} as our nested sampler. Nested sampling, an algorithm introduced by~\citet{skilling2004, skilling2006}, provides an estimate of the true Bayesian evidence and is often utilised for parameter estimation within the LIGO collaboration~\cite{bilby, bilby_paper, pbilby_paper}.

The most computationally intensive step during Bayesian inference is evaluating the likelihood $\mathcal{L}(d_i|\template)$. To accelerate our analysis, we use a likelihood that explicitly marginalises over coalescence time, phase at coalescence, and luminosity distance (Eq.~80 from~\citet{intro_to_gw_bayes}). While this marginalised likelihood reduces the run time without introducing errors to our evidence evaluation, it does not generates samples for the marginalised parameters. However, these parameter samples can be calculated as a post-processing step~\cite{intro_to_gw_bayes}.

@@

The waveform template we utilise is \imrphenomp, a phenomenological waveform template constructed in the frequency domain that models the in-spiral, merger, and ringdown (IMR) of a compact binary coalescence~\citep{khan2016frequency}. Although there exist gravitational-wave templates such as \seob~\cite{seobnrv4phm} which incorporate more physics, such as information on higher-order modes, we use \imrphenomp as it is computationally inexpensive compared to others.

We take 31 neighbouring, off-source, non-overlapping,  4-second  segments of time-series data before the analysis data segment $d_i$ to generate the PSD. We use off-source to avoid the inclusion of a signal in the PSD calculation. A Tukey window with 0.2-second roll-offs is applied to each data segment to suppress spectral leakage after which the segments are fast-Fourier transformed and median-averaged to create a PSD~\cite{ligo_psd}. Like other PSD estimation methods, this method adds statistical uncertainties to the PSD~\cite{psd_student_t, chatziioannou2019noise}. To marginalise over the statistical uncertainty, we use the median-likelihood presented by~\citet{psd_student_t} as a post-processing step. 
\avi{Greg would like me to discuss the effect of the PSD marginalization in more detail}

@@

Using the tuned prior odds the \bcr for the candidate events can be calculated. Figure~\ref{fig:bcrCdf} shows the \bcr distributions for the background triggers, simulated triggers and candidate events. The bulk of the background and simulated trigger distributions are separate but slightly overlap due to some of the simulated signal's being very faint. The separation suggests that the \bcr can successfully distinguish signals from noise or glitches. The vertical lines in Figure~\ref{fig:bcrCdf} displays the \bcr for gravitational-wave candidate events. On comparing the candidate event \bcr values with the background distribution, we can estimate \pastro values for the candidate events.

\begin{figure}[!ht]
{\centering \includegraphics[width=0.85\linewidth]{images/bcr_cdf_smaller_legend.png} }
\caption[BCR distribution example]{Histograms represent the survival function (1-CDF) from our high-mass selection of background triggers (grey) and simulated signals (blue) triggers obtained from \pycbc's search of data from $\text{August 13 - 21, 2017}$. Vertical lines mark the $\text{ln}\ \bcr$ of IAS's GW170817 and GWTC-1's GW170814.}\label{fig:bcrCdf}
\end{figure}

@@

\input{Data/results_table}
We analyse the $60\ 996$ background, $5\ 146$ simulated, and $25$ candidate triggers reported by \pycbc's search on the data from LIGO's second observing run, restricting our analysis to the triggers that fall within our mass-space as described in Section~\ref{sec:method}. In addition to these triggers, we also analyse events and candidate events reported by GWTC-1 and the IAS group (note that some of these were identified as candidates by the \pycbc search). In Table~\ref{tab:results}, we summarise the \pastrobcr, along with the \pastro of other pipelines for comparison. Although the various pipeline \pastro are not mathematically equivalent, by comparing pipeline \pastro values for a given candidate, we can compare how significant each pipeline deems various candidates. The $\alpha$ and $\beta$ values utilised for each time-frame are reported in Appendix~\ref{apdx:alphabeta}.

@@

>>>>

One proposed heave-seed pathway involves the direct collapse of massive, metal-free Population III stars. Typically, massive stars $> 100 \ \msun$ can eject more than half of their mass in the form of radiatively driven stellar winds~\cite{}. However, at zero metallicity, mass loss through stellar winds is negligible \cite{Kudritzki 2000}. Consequently,  Population III stars have the opportunity to complete their life-cycles without losing a significant portion of their mass. Moreover, Population III stars with masses above $230 \ \msun$ will avoid electron-positron pair-instability. Therefore, oxygen and silicon burning will not drive an explosion, allowing the star to undergo a complete collapse to leave behind an intermediate-mass black hole remnant with a mass of $120-280 \ \msun$, dependent on the initial star mass \cite{Bond et al. 1984}. As the fraction of Population III stars with masses above $230 \ \msun$ is uncertain, the number of expected intermediate-mass black holes from these stars is also ambiguous~\cite{Coleman Miller, Formation Scenarios for Intermediate-Mass Black Holes}.

@@

\rs{\sout{It is possible to measure the mass of nearby supermassive black holes by directly observing the dynamic motion of stars nearby the supermassive black hole (such as Sgr A$^*$~). However, this dynamic measurement method is challenging to utilise for  intermediate-mass black holes as their sphere of influence is too small to resolve motions~\cite{}. Additionally, dynamic-measurements that have alluded to  intermediate-mass candidates, e.g. the candidate at the center of the globular cluster 47 Tucanae~\cite{kiziltan 2017}, can also be attributed to simpler models without a central  intermediate-mass black hole~\cite{freire 2017}. }}

@@

Although a significant fraction of the glitches can be removed by testing them for coherence amongst various ground based detectors and performing matched-filtering, these methods are insufficient to remove all the glitches. One method to account for more glitches while searching for high total mass CBC gravitational-waves is by utilizing the astrophysical Bayesian odds~\cite{bci, BCR1, BCR2, bcr_gw151216, bayesian_odds}. A true Bayesian odds, calculated without using bootstrap techniques, can provide more accurate estimate of significance that does note depend on the search pipeline~\cite{BCR2, bcr_gw151216,  bayesian_odds}. In this paper, we utilize a Bayesian method, called the Bayesian Coherence Ratio \bcr~\cite{BCR1}, to rank the candidate gravitational-wave signals from high-mass compact binary coalescences (systems with total masses in the range of $55-500\ \msun$) in the detector data recorded during O2. Although the \bcr, utilizing bootstrap techniques, does not provide the true Bayesian odds,  it utilizes Bayesian evidences which describe the explicit probability of data under the hypothesis that it contains coherent signals vs incoherent glitches.

@@

The standard framework to identify CBC gravitational-wave signals hidden in data is by quantifying the significance of candidates with null-hypothesis significance testing. In this framework, the candidates' ranking statistic is compared against a background distribution in a frequentist approach. \rs{add a sentence describing what the ranking stats are for the different pipelines (key words: matched filtering, SNR etc...) and add citations!} On the other hand, the standard framework for performing parameter estimation and model selection in gravitational-wave astronomy is Bayesian inference. This work utilizes Bayesian inference to calculate the Bayesian Coherence odds-ratio~\cite{BCR1}, \bcr, of high-mass candidates in LIGO's second observing run. We use the \bcr not as an odds-ratio but instead as a ranking statistic, a step toward building a unified Bayesian framework to search for candidates and estimate their parameters.

@@

Bayes theorem states that the posterior probability distribution $p(\parameters|d,\mathcal{H})$ for data $d$ and a vector of parameters \parameters that describe a model which quantifies a hypothesis $\mathcal{H}$, is given by
\begin{equation}
p(\parameters|d,\mathcal{H}) = \frac{\mathcal{L}(d|\parameters, \mathcal{H}) \ \pi(\parameters | \mathcal{H})}{\mathcal{Z}(d|\mathcal{H})}\ , 
\end{equation}
where $\mathcal{L}(d|\parameters, \mathcal{H})$ is the likelihood of the data given the parameters \parameters and the hypothesis, $\pi(\parameters | \mathcal{H})$ is the prior probability of the parameters, and finally,
\begin{equation}
    \mathcal{Z}(d|\mathcal{H}) = \int\limits_{\parameters} \ \mathcal{L}(d|\parameters,\mathcal{H}) \ \pi(\parameters | \mathcal{H}) \ d\parameters
\end{equation} is the likelihood after marginalizing over the parameters \parameters.  To compare two hypotheses $\mathcal{H}_A$ and $\mathcal{H}_B$ with the Bayes theorem one can calculate an odds-ratio
\begin{equation}
    \mathcal{O}^A_B = \frac{\mathcal{Z}^A\ \pi(\parameters^A)}{\mathcal{Z}^B\ \pi(\parameters^B)}\ ,
\end{equation}
where $\mathcal{Z}^A$ and $\mathcal{Z}^B$ are the shorthand for the evidences  $\mathcal{Z}(d|\mathcal{H}_A)$ and $\mathcal{Z}(d|\mathcal{H}_B)$. The odds-ratio can tell us which of the two hypotheses is more likely. For example, if $\mathcal{O}^A_B >> 1$, then this odds ratio indicates that the $\mathcal{H}_A$ describes the data much better than $\mathcal{H}_B$.

@@

For our study, we filter the background, simulated and candidate events to include only high-mass events with masses in the ranges of the parameters presented in Table~\ref{tab:parameters}. A plot of the \pycbc triggers from one time-frame, during April 23 - May 8, 2017, is presented in Fig.~\ref{fig:templateBank}. This figure also depicts the gravitational-wave templates used during the search through this time-frame of data.

@@

\subsection{Calculating the BCR for triggers}
\avi{Sat 13th: Just realised that some parts of this dont make sense without the material present in Apdx~\ref{sec:bayesianEvidEval}}...
To evaluate $Z^S$, $Z^G_i$ and $Z^N_i$ and calculate the \bcr Eq.~\ref{eq:bcr} for triggers, we carry out Bayesian inference with \bilby~\cite{bilby, bilby_pipe}, employing \dynesty~\cite{dynesty} as our nested sampler. Nested sampling, an algorithm introduced by~\citet{skilling2004, skilling2006}, provides an estimate of the true Bayesian evidence and is often utilized for parameter estimation within the LIGO collaboration~\cite{bilby, bilby_paper, pbilby_paper}.

The most computationally intensive step during Bayesian inference is evaluating the likelihood $\mathcal{L}(d_i|\template)$. To accelerate our analysis, we use a likelihood that explicitly marginalizes over coalescence time, phase at coalescence, and luminosity distance (Eq.~80 from~\citet{intro_to_gw_bayes}). While this marginalized likelihood reduces the run time without introducing errors to our evidence evaluation, it does not generates samples for the marginalized parameters. However, these parameter samples can be calculated as a post-processing step~\cite{intro_to_gw_bayes}.

@@

The waveform template we utilize is \imrphenomp, a phenomenological waveform template constructed in the frequency domain that models the in-spiral, merger, and ring-down (IMR) of a compact binary coalescence~\citep{khan2016frequency}. Although there exist gravitational-wave templates such as \seob~\cite{seobnrv4phm} which incorporate more physics, such as information on higher-order modes, we use \imrphenomp as it is computationally inexpensive compared to others.

We take 31 neighboring, off-source, non-overlapping,  4-second  segments of time-series data before the analysis data segment $d_i$ to generate the PSD. We use off-source to avoid the inclusion of a signal in the PSD calculation. A Tukey window with 0.2-second roll-offs is applied to each data segment to suppress spectral leakage after which the segments are fast-Fourier transformed and median-averaged to create a PSD~\cite{ligo_psd}. Like other PSD estimation methods, this method adds statistical uncertainties to the PSD~\cite{psd_student_t, chatziioannou2019noise}. To marginalize over the statistical uncertainty, we use the median-likelihood presented by~\citet{psd_student_t} as a post-processing step. 
\avi{Greg would like me to discuss the effect of the PSD marginalization in more detail}

We take 31 neighboring off-source non-overlapping  4-second  segments of time-series data before the analysis data segment $d_i$ to generate the PSD.  A Tukey window with a 0.2-second roll-off is applied to each data segment to suppress spectral leakage. After this  the segments are fast-Fourier transformed and median-averaged to create a PSD~\cite{ligo_psd}. Like other PSD estimation methods, this method adds statistical uncertainties to the PSD~\cite{psd_student_t, chatziioannou2019noise, Biscoveanu:2020:PhRvD}. To marginalize over the statistical uncertainty, we use the median-likelihood presented by~\citet{psd_student_t} as a post-processing step. We find that this post-processing step improves the search efficiency by $x\%$ the details of this calculation are presented in the Appendix~\ref{sec:psd-marginalization}.

@@

Using the tuned prior odds the \bcr for the candidate events can be calculated. Fig.~\ref{fig:bcrCdf} shows the \bcr distributions for the background triggers, simulated triggers and candidate events. The bulk of the background and simulated trigger distributions are separate but slightly overlap due to some of the simulated signal's being very faint. The separation suggests that the \bcr can successfully distinguish signals from noise or glitches. The vertical lines in Fig.~\ref{fig:bcrCdf} displays the \bcr for gravitational-wave candidate events. On comparing the candidate event \bcr values with the background distribution, we can estimate \pastro values for the candidate events.

\begin{figure}[!ht]
{\centering \includegraphics[width=0.85\linewidth]{images/bcr_cdf_smaller_legend.png} }
\caption[BCR distribution example]{Histograms represent the survival function (1-CDF) from our high-mass selection of background triggers (grey) and simulated signals (blue) triggers obtained from \pycbc's search of data from $\text{August 13 - 21, 2017}$. Vertical lines mark the $\text{ln}\ \bcr$ of IAS's GW170817A and GWTC-1's GW170814.}\label{fig:bcrCdf}
\end{figure}

@@

\input{Data/results_table}
We analyse the $60,996$ background, $5,146$ simulated, and $25$ candidate triggers reported by \pycbc's search on the data from LIGO's second observing run, restricting our analysis to the triggers that fall within our mass-space as described in Section~\ref{sec:method}.We also analyse events and candidate events reported by GWTC-1 and the IAS group (note: some of these were identified as candidates by the \pycbc search). In Table~\ref{tab:results}, we summarise the \pastrobcr, along with the \pastro of other pipelines for comparison. Although the various pipeline \pastro are not mathematically equivalent, by comparing pipeline \pastro values for a given candidate, we can compare how significant each pipeline deems various candidates. The $\alpha$ and $\beta$ values utilised for each time-frame are reported in Appendix~\ref{apdx:alphabeta}.

@@

\section{Marginalizing over PSD statistical uncertainties}\label{sec:psd-marginalization}
To generate the results in Fig.~\ref{fig:bcrCdf}, we applied a post-processing step to marginalize the uncertainty in the PSD. In Fig.~\ref{fig:bcrCdfUnmarginalized}, we show the results if this post-processing step is not applied. Clearly, marginalizing over uncertainty in the PSD yields an improvement in the separation of the noise and signal distributions. Quantitatively, this results in a reduction in the rate of false alarms from X to Y and a reduction in the rate of false positives from X to Y in the August 13 - 21, 2017 time-frame of data. For the entirety of O2 PSD marginalisation resulted in a reduction of X, or an $x\%$ improvement in search efficency.

\begin{figure}[!ht]
{\centering \includegraphics[width=0.85\linewidth]{images/bcr_cdf_smaller_legend.png} }
\caption[BCR distribution example]{This plot is analogous to Fig.~\ref{fig:bcrCdf}, but without using the post-processing step to marginalize over PSD statistical uncertainties. Without the post-processing step, there is a greater overlap between the background (grey) and foreground (blue) survival functions. For more details about this plot, refer to the caption of Fig.~\ref{fig:bcrCdf}.}\label{fig:bcrCdfUnmarginalized}
\end{figure}
